---
title: "3.1: Voting I: Preference Aggregation & Cycling"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

# Appendix: Spatial Competition in Multiple-Dimensions

In class, we looked at spatial competition along a single dimension, and saw that **if preferences are single-peaked**, the **median** platform will win a majority, and escape the problem of cycling.

Suppose we have three voters again voting on a platform of **two** issues: (1) Education Spending and (2) Defense Spending.


```{r, echo=F, warning=F, message=F}
library("tidyverse")
library("ggrepel")
library("ggforce")

```

```{r, fig.retina=3, echo=F}
points<-tribble(
  ~x, ~y, ~label,
  25, 75, "A",
  75, 50, "B",
  50, 25, "C"
)

base<-ggplot(data = points)+
  aes(x = x,
      y = y)+
  geom_point(size=3, aes(color = label))+
  geom_label_repel(aes(color = label,
                       label = label), seed=256, size = 6)+
    labs(x = "Education Spending (Billions)",
       y = "Defense Spending (Billions)")+
  scale_x_continuous(breaks = seq(0,100,10),
                     limits = c(0,100),
                     expand = c(0,0),
                     labels = scales::dollar
                     )+
  scale_y_continuous(breaks = seq(0,100,10),
                     limits = c(0,100),
                     expand = c(0,0),
                     labels = scales::dollar
                     )+
  guides(color=F)+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
base
```

We represent their **ideal points** in 2-dimensional issue space in the figure above. Imagine this is looking down on the peaks of three mountains, as moving in *any direction* away from a voter's ideal point (peak) decreases their utility (if you like, decreases in elevation from the peak).

Now, we can still apply the median voter theorem here, by determining the median of each issue. For Education Spending, C is the median (A to the left and C to the right), and for Defense Spending, B is the median (A above, and C below). 

```{r, fig.retina=3, echo=F}
base+
  geom_vline(xintercept = 50, size = 1, linetype="dashed")+
  geom_hline(yintercept = 50, size = 1, linetype="dashed")+
  geom_point(x = 50, y = 50, size =3, color = "purple")+
  geom_label_repel(data = tibble(x=50,
                                 y=50),
                   aes(x=x,
                       y=y),
                   label = "Median",
                   color = "purple",
                   seed=256)
```

We have found the median platform by combining the median of each issue. Now, find the distance, for each voter, from their ideal point to the median platform. Consider this a distance the radius of a circle. This circle, for each voter, represents an **indifference curve** for that voter: the set of alternative platforms of education and defense that is *equally preferable* (gives same utility) for that voter as the median platform.

*Inside* each circle is the set of platforms that are **preferred over** the median (and equivalent platforms, the circumference of the circle). As an obvious demonstration, each voter prefers their own ideal point (A prefers A, B prefers B, and C prefers C) over the median platform.

```{r, fig.retina=3, echo=F}
lines<-tribble(
  ~x, ~xend, ~y, ~yend, ~label,
  25, 50, 75, 50, "A",
  75, 50, 50, 50, "B",
  50, 50, 25, 50, "C"
)

circles<-tribble(
  ~x, ~y, ~radius, ~label,
  25, 75, 35, "A",
  75, 50, 25, "B",
  50, 25, 25, "C"
)


ggplot(data = points)+
  aes(x = x,
      y = y)+
  geom_point(size=3, aes(color = label))+
  geom_label_repel(aes(color = label,
                       label = label), seed=256, size = 6)+
    labs(x = "Education Spending (Billions)",
       y = "Defense Spending (Billions)")+
  geom_point(x = 50, y = 50, size =3, color = "purple")+
  geom_circle(data = circles,
              aes(x0 = x,
                  y0 = y,
                  r = radius,
                  color = label,
                  fill = label),
              inherit.aes = F,
              alpha = 0.4,
              size = 1)+
  geom_segment(data = lines, aes(x=x,
                                 xend=xend,
                                 y=y,
                                 yend=yend,
                                 color=label),
               size = 1.5,
               linetype="dashed")+
  geom_label_repel(data = tibble(x=50,
                                 y=50),
                   aes(x=x,
                       y=y),
                   label = "Median",
                   color = "purple",
                   seed=256)+
  guides(fill = F, color = F)+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)+
    scale_x_continuous(breaks = seq(0,100,10),
                     #limits = c(0,100),
                     expand = c(0,0),
                     labels = scales::dollar
                     )+
  scale_y_continuous(breaks = seq(0,100,10),
                     #limits = c(0,100),
                     expand = c(0,0),
                     labels = scales::dollar
                     )+
  coord_fixed(clip="off")

```

Note the "preferred-to sets" (the shaded circles) **do not overlap** at the median! The borders of the circles do, i.e. the indifference curve, but not the preferred-to sets! 

Now, allow the voters to propose any platform they wish, to vote on in a pairwise vote. We begin with the median as a starting point: ($50B Education Spending, $50B Defense Spending).

Consider voter $A$. She could propose something she prefers over the median, say point $1$ at ($30, $70), but that is outside the other voters' preferred-to sets (outside of their circles), so this would be **worse** for them, and they would vote against it. 

```{r, fig.retina=3, echo=F}

votes<-ggplot(data = points)+
  aes(x = x,
      y = y)+
  geom_point(size=3, aes(color = label))+
  geom_label_repel(aes(color = label,
                       label = label), seed=256, size = 6)+
    labs(x = "Education Spending (Billions)",
       y = "Defense Spending (Billions)")+
  geom_point(x = 50, y = 50, size =3, color = "purple")+
  geom_circle(data = circles,
              aes(x0 = x,
                  y0 = y,
                  r = radius,
                  color = label,
                  fill = label),
              inherit.aes = F,
              alpha = 0.4,
              size = 1)+
  geom_label_repel(data = tibble(x=50,
                                 y=50),
                   aes(x=x,
                       y=y),
                   label = "Median",
                   color = "purple",
                   seed=256)+
  guides(fill = F, color = F)+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)+
    scale_x_continuous(breaks = seq(0,100,10),
                     #limits = c(0,100),
                     expand = c(0,0),
                     labels = scales::dollar
                     )+
  scale_y_continuous(breaks = seq(0,100,10),
                     #limits = c(0,100),
                     expand = c(0,0),
                     labels = scales::dollar
                     )+
  coord_fixed(clip="off")

votes+
  geom_point(x = 30, y = 70, size = 2)+
  geom_label_repel(data = tibble(x=30,
                                 y=70),
                   aes(x=x,
                       y=y),
                   label = "1",
                   color = "black",
                   seed=256)
```

If $A$ proposes something inside her and any other voter's preferred-to set (inside the circle), she can get a majority to approve it over the median. Suppose she proposes platform $D$: $35B education and $42.5 B defense. Then her and voter $C$ will vote for this over the median, since it is inside both of their circles (and thus, preferred over the median), whereas $B$ will vote against it, it being worse than the median. 

```{r, fig.retina=3}
votes+
  annotate("segment", x = 50, xend = 35, y = 50, yend = 42.5, color = "purple", size=1.5, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  geom_point(x = 35, y = 42.5, size = 2)+
  geom_label_repel(data = tibble(x=35,
                                 y=42.5),
                   aes(x=x,
                       y=y),
                   label = "D",
                   color = "black",
                   seed=256)

```
Platform $D$ wins, 2-1. Suppose voter $B$ suggests an alternative to $D$, platform $E$, with $58B in Education and $33B in Defense. Compared to platform $D$, voters $B$ and $C$ prefer this, as it is closer to their ideal points, but it is well outside voter $A$'s circle, so she votes against. But the vote of $E$ vs $D$ results in $E$ winning, 2-1. 

```{r, fig.retina=3}
votes+
  annotate("segment", x = 50, xend = 35, y = 50, yend = 42.5, color = "purple", size=1.5, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  
  annotate("segment", x = 35, xend = 58, y = 42.5, yend = 33, color = "purple", size=1.5, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
# point D
    geom_point(x = 35, y = 42.5, size = 2)+
  geom_label_repel(data = tibble(x=35,
                                 y=42.5),
                   aes(x=x,
                       y=y),
                   label = "D",
                   color = "black",
                   seed=256)+
# point E
      geom_point(x = 58, y = 33, size = 2)+
  geom_label_repel(data = tibble(x=58,
                                 y=33),
                   aes(x=x,
                       y=y),
                   label = "E",
                   color = "black",
                   seed=256)


```

Suppose $A$ counters proposal $E$ with a new proposal, $F$: $55B in Education, $58B in Defense. Now $A$ and $B$ prefer this to proposal $D$, while $B$ votes against it. Measure $F$ passes, 2-1: 

```{r, fig.retina=3}
votes+
  annotate("segment", x = 50, xend = 35, y = 50, yend = 42.5, color = "purple", size=1.5, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  
  annotate("segment", x = 35, xend = 58, y = 42.5, yend = 33, color = "purple", size=1.5, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  
  annotate("segment", x = 58, xend = 55, y = 33, yend = 58, color = "purple", size=1.5, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+

# point D
    geom_point(x = 35, y = 42.5, size = 2)+
  geom_label_repel(data = tibble(x=35,
                                 y=42.5),
                   aes(x=x,
                       y=y),
                   label = "D",
                   color = "black",
                   seed=256)+
# point E
      geom_point(x = 58, y = 33, size = 2)+
  geom_label_repel(data = tibble(x=58,
                                 y=33),
                   aes(x=x,
                       y=y),
                   label = "E",
                   color = "black",
                   seed=256)+
# point F
      geom_point(x = 55, y = 58, size = 2)+
  geom_label_repel(data = tibble(x=55,
                                 y=58),
                   aes(x=x,
                       y=y),
                   label = "F",
                   color = "black",
                   seed=256)

```

Now we see that, in multi-issue space, **we get cycling**! Any voter can propose a platform that can capture a majority^[On the graph, place the platform inside an intersection between two voters' circles. Those two voters vote for, the other voter votes against.], and any voter can then propose an alternative platform that can also capture a majority for the alternative platform and against the first platform!

This is sometimes called the **Chaos Theorem** in social choice theory: in multi-dimensional policy space, even with single-peaked preferences^[Remember, the peak is the ideal point for each voter, utility decreases with distance from that point in all directions.], there are no Condorcet winners, and thus a majority opposes each alternative. We can reach **any** point in this space with a carefully-constructed sequence of votes. Thus, again, there is no "will of the majority" and an agenda-setter can achieve any outcome they want.

The Median Voter Theorem worked in a single dimension with single-peaked preferences. The result we see here, in multiple issue space (and generalized by Arrow's impossibility theorem), is that cycling is not easily solved: the median voter theorem fails when there are multiple issues.^[Or conversely, the median voter theorem only works for single-peaked preferences for a *single* issue.]